---
title: "sf_traits_analysis"
author: "Liyenne"
date: "2023-11-09"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("outputs")
    )
  })
cache: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```

# setup
```{r packages}
require(readr)
require(tidyverse)

require(vegan)
require(lme4)
require(lsmeans)
require(traitstrap)

require(ggplot2)
require(ggpubr)
require(ggridges)
require(GGally)

require(ggplot2)
require(ggrepel)
require(ggtext)
require(ggpubr)
require(ggordiplots)
require(ggvegan)
require(ggsignif)
```

# Data import
```{r}
# taxonomic information
species_list <- read_csv("2_data/1_raw_data/community_composition/sf_vp_species_list.csv") %>% 
  mutate(across(1:5, as.factor))
```

```{r data import}
# trait data
trait_data_raw <- read_csv("2_data/1_raw_data/snowfence_traits/sf_plant_trait_data_full_v1.csv") %>% 
  mutate(habitat = case_when(habitat == 'D' ~ 'fen',
                           habitat == 'E' ~ 'heath',
                           habitat == "X" ~ 'ridge')) %>% 
  mutate(treatment = case_when(treatment == 'C' ~ 'control',
                           treatment == 'SF' ~ 'snowfence')) %>% 
  left_join(species_list, by = "spec") %>% # add taxonomic data
  select(code:spec, species:group, everything())
```

```{r}
# community composition data (long)
community_data <- read_csv("2_data/1_raw_data/community_composition/vp_pp_leaf_2025_l_fn.csv") %>% 
  mutate(habitat = as.factor(ifelse( # create habitat variable based on plot code
    grepl('D', plot), 'fen', ifelse(
      grepl('E', plot), 'heath', 'ridge'))
    )) %>% 
  mutate(treatment = as.factor(ifelse( # create treatment variable based on plot code
    grepl('C', plot), 'control', 'snowfence'
    ))) %>%  
  mutate(site = as.factor(substr(plot, 2,2))) %>% # create site variable from plot code
  left_join(species_list, by = "species") %>% 
  select(plot, habitat, site, treatment, spec, species, family:group, hits)
```

```{r}
leaf_chem <- read_csv("2_data/1_raw_data/snowfence_traits/sf_leaf_chem.csv") %>%
  mutate(across(1:3, as.factor))
```


# Wrangling
## trait data (full)
```{r LDMC caclulation}
trait_data <- trait_data_raw %>% 
  mutate(LDMC = dweight / fweight) %>% 
  select(code:leafa, LDMC); head(trait_data, 10)
```

```{r SLA caclulation}
trait_data <- trait_data_raw %>% 
  mutate(SLA = leafa / dweight) %>% 
  full_join(trait_data) %>% 
  select(code:dweight, leafa, LDMC, SLA); head(trait_data, 10)
```

```{r}
trait_data <- trait_data %>% 
  mutate(across(1:10, as.factor)) # convert catergorical variables to factor
```

```{r}
# convert trait dataset to long format
trait_data_long <- trait_data %>% 
  pivot_longer(!code:group, names_to = "trait", values_to = "trait_value") 
```

## trait data (pooled) 
```{r}
trait_data_pooled <- trait_data %>%  # pool data by plot and species
  group_by(plot, spec) %>% 
  summarise(height = mean(height),
            dweight = mean(dweight), 
            fweight = mean(fweight),
            leafa = mean(leafa), 
            LDMC = mean(LDMC), 
            SLA = mean(SLA)) 

trait_data_pooled <- trait_data %>% # add taxonomic variables to pooled dataset
  select(2:10) %>% 
  right_join(trait_data_pooled, by = c("plot", "spec")) %>% 
  distinct()
```

### add leaf chem comp
```{r}
trait_data_pooled <- trait_data_pooled %>% 
  left_join(leaf_chem, by = c("plot", "spec")) %>% # combined data has 3 extra rows??? 
  select(ID, everything()) %>% 
  mutate(across(c(wN:CN_ratio), as.numeric))
```

```{r}
# convert trait dataset to long format
trait_data_pooled_long <- trait_data_pooled %>% 
  pivot_longer(!ID:group, names_to = "trait", values_to = "trait_value") %>% 
  filter(!is.na(ID))
```

## analytics
```{r}
# in how many plots did each species occur? 
x <- trait_data_raw %>% 
  group_by(spec) %>% 
  summarise(unique_plot_count = n_distinct(plot)) %>% 
  arrange(desc(unique_plot_count)); print(x)

# how many species occurred in 10 or more plots
x %>% 
  filter(unique_plot_count >= 10) %>% 
  nrow() # 12

# how many pooled samples does that constitute?
x %>% 
  filter(unique_plot_count >= 10) %>% 
  summarise(n_pools = sum(unique_plot_count))
```

# Bootstrapping traits
## trait fill 
### no chem
```{r}
filled_traits <- trait_fill(
comm = community_data, traits = trait_data_long,
scale_hierarchy = c("habitat", "site", "treatment", "plot"),
taxon_col = c("species", "genus", "family"), value_col = "trait_value",
trait_col = "trait", abundance_col = "hits", complete_only = TRUE, leaf_id = "code"
)
autoplot(filled_traits)
```

```{r}
trait_missing(filled_trait = filled_traits, comm = community_data)
```

```{r}
boot_traits <- trait_np_bootstrap(filled_traits = filled_traits, nrep = 10, sample_size = 100)
```

### with chem
```{r}
filled_traits_pooled <- trait_fill(
comm = community_data, traits = trait_data_pooled_long,
scale_hierarchy = c("habitat", "site", "treatment", "plot"),
taxon_col = c("species", "genus", "family"), value_col = "trait_value",
trait_col = "trait", abundance_col = "hits", complete_only = TRUE, leaf_id = "ID"
)
autoplot(filled_traits_pooled)
```

```{r}
trait_missing(filled_trait = filled_traits_pooled, comm = community_data)
```

```{r}
boot_traits_pooled <- trait_np_bootstrap(filled_traits = filled_traits_pooled, nrep = 10, sample_size = 100)
```

# Data exploration
```{r}
trait_data_raw %>% 
  ggplot(aes(x = fweight, y = dweight, color = treatment)) +
  geom_point() +
  theme_classic() 
```



## density plots (unweighted)
```{r}
# density plots for plant height
trait_data %>% 
  ggplot(aes(x = height, y = habitat, fill = habitat)) +
  geom_density_ridges()

trait_data %>% 
  ggplot(aes(x = height, y = treatment, fill = treatment)) +
  geom_density_ridges()
```

```{r}
# density plots for LDMC
trait_data %>% 
  ggplot(aes(x = LDMC, y = habitat, fill = habitat)) +
  geom_density_ridges()

trait_data %>% 
  ggplot(aes(x = LDMC, y = treatment, fill = treatment)) +
  geom_density_ridges()
```

```{r}
# density plots for SLA
trait_data %>% 
  ggplot(aes(x = SLA, y = habitat, fill = habitat)) +
  geom_density_ridges()

trait_data %>% 
  ggplot(aes(x = SLA, y = treatment, fill = treatment)) +
  geom_density_ridges()
```

## per species
How does individual species response differ between treatments? what is the direction of this response and what is it related to? 
```{r plant height plot, echo = FALSE}
trait_data %>% 
  ggplot(aes(x = habitat, y = height, color = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~ spec) +
  theme(legend.position = c(.85, .8), 
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("3_output/Figures/plant_height.svg", height = 4, width = 25)
```

```{r LDMC per species, echo = FALSE}
trait_data %>% 
  ggplot(aes(x = habitat, y = LDMC, color = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~ spec) +
  theme(legend.position = c(.85, .8), 
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("3_output/Figures/LDMC.svg", height = 4, width = 25)
```

```{r SLA per species plot, echo = FALSE}
trait_data %>% 
  ggplot(aes(x = habitat, y = SLA, color = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~ spec) +
  theme(legend.position = c(.85, .8), 
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("3_output/Figures/SLA.svg", height = 4, width = 25)
```

```{r}
trait_data_pooled %>% 
  ggplot(aes(x = habitat, y = CN_ratio, color = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~ spec) +
  theme(legend.position = c(.85, .8), 
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("3_output/Figures/CN_ratio.svg", height = 4, width = 25)
```


```{r}
filled_traits_pooled %>% 
  filter(trait == "height") %>% 
  ggplot(aes(x = habitat, y = trait_value, color = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~ spec_comm) +
  theme(legend.position = c(.85, .8), 
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("3_output/Figures/height_filled.svg", height = 4, width = 25)
```


## community weighted averages
are shifts in community averages driven by species adaptation or species composition differences?
### no chem
```{r}
boot_traits %>% 
  ggplot(aes(x = habitat, y = mean, fill = treatment)) + 
  geom_boxplot() + 
  facet_wrap(~ trait, scales = "free") + 
  theme_classic() +
  theme(panel.spacing = unit(.0, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "bottom",
        legend.margin = margin(t = -8),
        strip.background = element_rect(color = "black", size = 1),
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("../3_output/Figures/bootstrapped_traits.svg", height = 7, width = 7)
```

### with chem
```{r}
facet_order <- c("height", "SLA", "LDMC", "CN_ratio", "d13C", "d15N")
trait_labs <- c("plant height", "SLA", "LDMC", "C:N ratio", as.character(bquote(δ^13*C)), as.character(bquote(δ^15*N)))
names(trait_labs) <- c("height", "SLA", "LDMC", "CN_ratio", "d13C", "d15N")
boot_traits_pooled <- boot_traits_pooled %>% mutate(habitat = recode(habitat,
                             "meadow" = "fen"))

boot_traits_pooled %>% 
  filter(trait %in% c("SLA", "LDMC", "height", "CN_ratio", "d13C", "d15N")) %>%
  mutate(trait = factor(trait, levels = facet_order)) %>%
  mutate(habitat = factor(habitat, levels = c("fen", "heath", "ridge"))) %>%
  ggplot(aes(x = habitat, y = mean, fill = treatment)) + 
  geom_boxplot() + 
  scale_fill_manual(values = c("#999999", "#4da6ff")) +
  facet_wrap(~ trait, scales = "free", ncol = 2, nrow = 3,
             labeller = labeller(trait = trait_labs)) + 
  theme_classic() +
  theme(panel.spacing = unit(.0, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "bottom",
        legend.margin = margin(t = -8),
        strip.background = element_rect(color = "black", size = 1),
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("3_output/Figures/bootstrapped_traits_pooled_2025.svg", height = 7, width = 3)
```

## strategies
```{r}
trait_data %>% 
  ggplot(aes(x = SLA, y = LDMC, colour = group)) +
  geom_point()
```

## local vs global

## PCA 
```{r}
trait_data_pooled %>% 
  #select(species, height:CN_ratio) %>% 
  na.omit() %>% 
  ggpairs(columns = 11:21, 
          mapping = aes(color = habitat))

ggsave("../3_output/Figures/variables.svg")
```

### species traits
```{r}
species_traits <- trait_data_pooled %>% 
  filter(complete.cases(.)) %>% 
  group_by(species) %>% 
  summarise(plant_height = mean(height), 
            LDMC = mean(LDMC), 
            SLA = mean(SLA), 
            N = mean(wN), 
            d15N = mean(d15N), 
            C = mean(wC), 
            d13C = mean(d13C), 
            CN_ratio = mean(CN_ratio)) %>% 
  column_to_rownames(var = "species")


```

```{r}
species_pca <- species_traits %>% 
  scale() %>% 
  rda()

summary(species_pca)

species_scores <- as.data.frame(species_pca$CA$u[, 1:2])
species_scores$species <- as.factor(rownames(species_scores))
trait_scores <- as.data.frame(species_pca$CA$v[, 1:2])
trait_scores$trait <- as.factor(rownames(trait_scores))

species_scores <- species_scores %>% 
  left_join(species_list)

```

```{r}
ggplot() +
  geom_hline(yintercept = 0, linetype = "dotted", alpha = 0.8) + 
  geom_vline(xintercept = 0, linetype = "dotted", alpha = 0.8) +
  geom_segment(data = trait_scores, aes(x = 0, y = 0, xend = 0.9 * PC1, yend = 0.9 * PC2), arrow = arrow(type = "closed", length = unit(0.1, "inches")), color = "gray") +
  geom_point(data = species_scores, aes(x = PC1, y = PC2, color = group), size = 3, alpha = 0.70) + # add the point markers
  geom_text_repel(data = species_scores, aes(x = PC1, y = PC2, label = species, color = group), size = 4,  alpha = 0.7) +  # add the species labels
  geom_label(data = trait_scores, aes(PC1, PC2, label = trait), col = "navy", label.size = NA, fill = NA) +
  #geom_text(data = data_scores, aes(x = NMDS1, y = NMDS2, label = plot), vjust=0) +  # add the site labels
  #coord_equal() +
  theme_classic() +
  #guides(color = guide_legend(override.aes = list(shape = NA))) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        strip.background = element_rect(color = "black", size = 1)) + 
  labs(x = "PC1 (50%)", y = "PC2 (17%)", color = "Functional group") 
  

ggsave("../3_output/Figures/species_pca.svg", height = 5, width = 6)
```

### trait observations
```{r}
pca_traits <- trait_data_pooled %>% 
  filter(complete.cases(.)) %>% 
  column_to_rownames(var = "ID") %>% 
  select(10, 14:20) %>% 
  scale() %>% 
  rda()

summary(pca_traits)

site_scores <- as.data.frame(pca_traits$CA$u[, 1:2])
site_scores$ID <- as.factor(rownames(site_scores))
trait_scores <- as.data.frame(pca_traits$CA$v[, 1:2])
trait_scores$trait <- as.factor(rownames(trait_scores))

site_scores <- trait_data_pooled %>%
  select(ID:group) %>% 
  right_join(site_scores)

site_scores[, 11:12] <- site_scores[, 11:12] * 2.5
```

```{r}
ggplot() +
  geom_hline(yintercept = 0, linetype = "dotted", alpha = 0.8) + 
  geom_vline(xintercept = 0, linetype = "dotted", alpha = 0.8) +
  geom_segment(data = trait_scores, aes(x = 0, y = 0, xend = 0.9 * PC1, yend = 0.9 * PC2), arrow = arrow(type = "closed", length = unit(0.1, "inches")), color = "gray") +
  geom_point(data = site_scores, aes(x = PC1, y = PC2, color = treatment), size = 3, alpha = 0.70) + # add the point markers
  geom_label(data = trait_scores, aes(PC1, PC2, label = trait), col = "navy", label.size = NA, fill = NA) +
  #geom_text(data = data_scores, aes(x = NMDS1, y = NMDS2, label = plot), vjust=0) +  # add the site labels
  #coord_equal() +
  theme_classic() +
  #guides(color = guide_legend(override.aes = list(shape = NA))) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        strip.background = element_rect(color = "black", size = 1)) + 
  labs(x = "PC1 (49%)", y = "PC2 (15%)") 
```

# Analysis
```{r}
# Betula nana
fit <- trait_data %>% 
  filter(species == "Betnan") %>% 
  lmer(height ~ habitat / treatment + (1 | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts

# Empetrum nigrum
fit <- trait_data %>% 
  filter(species == "Empnig") %>% 
  lmer(height ~ habitat / treatment + (1 | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts

# Vaccinium uliginosum
fit <- trait_data %>% 
  filter(species == "Vaculi") %>% 
  lmer(height ~ habitat / treatment + (habitat | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts

```

## Leaf Dry Matter Content (LDMC)

```{r LDMC plot, echo = FALSE}
trait_data %>% 
  ggplot(aes(x = habitat, y = LDMC, color = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~ species) +
  theme(legend.position = c(.85, .8))

#ggsave("~/Project Green Arctic/3_snow_fences/1_data/1_sf_data_analysis/3_output/3_plots/LDMC_23.svg", height = 4, width = 7)
```

```{r}
# Betula nana
fit <- trait_data %>% 
  filter(species == "Betnan") %>% 
  lmer(LDMC ~ habitat / treatment + (habitat | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts

# Empetrum nigrum
fit <- trait_data %>% 
  filter(species == "Empnig") %>% 
  lmer(LDMC ~ habitat / treatment + (habitat | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts

# Vaccinium uliginosum
fit <- trait_data %>% 
  filter(species == "Vaculi") %>% 
  lmer(LDMC ~ habitat / treatment + (habitat | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts
```

##Specific Leaf Area (SLA)

```{r SLA plot, echo = FALSE}
trait_data %>% 
  ggplot(aes(x = habitat, y = SLA, color = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~ species) 

ggsave("~/Project Green Arctic/3_snow_fences/1_data/1_sf_data_analysis/3_output/3_plots/SLA_23.svg", height = 4, width = 7)
```

```{r}
# Betula nana
fit <- trait_data %>% 
  filter(species == "Betnan") %>% 
  lmer(SLA ~ habitat / treatment + (habitat | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts

# Empetrum nigrum
fit <- trait_data %>% 
  filter(species == "Empnig") %>% 
  lmer(SLA ~ habitat / treatment + (habitat | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts

# Vaccinium uliginosum
fit <- trait_data %>% 
  filter(species == "Vaculi") %>% 
  lmer(SLA ~ habitat / treatment + (habitat | site), data = .)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ habitat / treatment, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts
```